---
title: "Level pool routing"
output: github_document
layout: post
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "../public/images/"
)
```

Level pool routing refers to one of the more straight-forward methods for calculating reservoir outflow given information about basin discharge relative to elevation and an input hydrograph (time vs inflow). Here reservoir is used in the technical engineering context and does not exclude the use of this method for natural lakes. I believe this method is applicable to any waterbody where you can assume a one-to-one relationship between discharge and elevation. Said another way, we assume that discharge does not have a hysteresis component and does not depend on the direction (rising vs falling limb) of the input hydrograph.

For more on this method see [here](http://www.engr.colostate.edu/~ramirez/ce_old/classes/cive322-Ramirez/CE322_Web/Example_LevelPoolRouting.htm), [here](https://www.caee.utexas.edu/prof/maidment/CE374KSpr12/Docs/Hmwk5Soln.pdf), and [here](https://www.caee.utexas.edu/prof/maidment/CE374KSpr13/Visual/HydrologicRouting.pptx). The following demonstration will layout the computational details and describe code to implement this method on three test datasets from the above links.

## Computation

```{r functions}

level_pool_routing <- function(lt, qh, area, delta_t, initial_outflow,
                               initial_storage, linear.fit){
  
  lagpad <- function(x, k) {
    c(rep(NA, k), x)[1 : length(x)] 
}
  
  lt$ii <- apply(cbind(lagpad(lt$inflow, 1), lt$inflow), 1, sum)

  if(is.null(qh$storage)){
    qh$storage <- area * qh$elevation
  }
  qh$stq     <- ((2 * qh$storage) / (delta_t)) + qh$discharge
  
  lt$sjtminq   <- NA
  lt$sj1tplusq <- NA
  lt$outflow   <- NA
  lt[1, c("sj1tplusq")] <- c(NA)
  lt[1, c("sjtminq")] <- ((2 * initial_storage) / delta_t) - initial_outflow
  
  lt[1, "outflow"] <- initial_outflow
  
  if(linear.fit == TRUE){
    fit <- lm(discharge ~ stq, data = qh)
  }else{
    fit <- mgcv::gam(discharge ~ s(stq, k = 2), data = qh)
    
    plot(qh$stq, qh$discharge)
    lines(qh$stq, predict(fit))
  }
  
  for(i in seq_len(nrow(lt))[-1]){
    lt[i, "sj1tplusq"] <- lt[i-1, "sjtminq"] + lt[i, "ii"]
    lt[i, "outflow"]   <- predict(fit, data.frame(stq = lt[i, "sj1tplusq"]))
    lt[i, "sjtminq"]   <- lt[i, "sj1tplusq"] - (lt[i, "outflow"] * 2)
  }

  lt
}

```


## Linear relationship between outflow and change in storage

The data for this example comes from this [level-pool routing walkthrough](http://www.engr.colostate.edu/~ramirez/ce_old/classes/cive322-Ramirez/CE322_Web/Example_LevelPoolRouting.htm). Here we are given storage as a function of discharge so we have no need for information on the area of the reservoir. Also, we are given an inflow hydrograph in 6 hour increments so we will need to specify a `delta_t` timestep in units of seconds.

```{r linear outflow-storage data}
input_hydro <- data.frame(
                time = seq(0, 162, by = 6),
                inflow = c(0, 50, 130, 250, 350, 540, 735, 1215, 1800, 1400,
                           1050, 900, 740, 620, 510, 420, 320, 270, 200, 150,
                           100, 72, 45, 25, 10, 0, 0, 0))

reservoir_char <- data.frame(elevation = c(130:134, 136:139),
                    discharge = c(20, 34, 57, 96, 162,
                    463, 781, 1318, 2226),
                    storage = c(1, 1.69, 2.85, 4.8, 8.12,
                    23.1, 39.1, 65.9, 111))
reservoir_char$storage <- reservoir_char$storage * 1000000
```

```{r linear outflow-storage exploratory plotting, eval=FALSE, echo=FALSE}
par(mfrow = c(1, 2))                                         
plot(reservoir_char$elevation, reservoir_char$storage, type = "b")
plot(reservoir_char$elevation, reservoir_char$discharge, type = "b")
par(mfrow = c(1, 1))
```

```{r linear outflow-storage calculation}
delta_t <- 6 * 3600

res_html <- level_pool_routing(input_hydro, reservoir_char, area = NA, delta_t = delta_t, initial_outflow = 20, initial_storage = 0, linear.fit = FALSE)
```

```{r linear outflow-storage final plotting}
plot(res_html$time, res_html$inflow)
lines(res_html$time, res_html$outflow)
```

[also this](https://www.caee.utexas.edu/prof/maidment/CE374KSpr12/Docs/Hmwk5Soln.pdf)

```{r }
library(pdftools)
txt  <- strsplit(pdf_text("https://www.caee.utexas.edu/prof/maidment/CE374KSpr12/Docs/Hmwk5Soln.pdf"), "\n")[[1]]

parse_table <- function(tbl, tbl_names){
  tbl <- lapply(tbl, function(x) read.table(text = x, stringsAsFactors = FALSE))
  tbl <- lapply(tbl, function(x) gsub(",", "", x))
  
  inds <- lapply(tbl, function(x) ifelse(max(grep(")", x)) == -Inf, 1, max(grep(")", x))))
  inds <- lapply(seq_along(tbl), function(x) c(1, (inds[[x]] + 1):length(tbl[[x]])))
  
  tbl <- lapply(seq_along(tbl), function(x) as.numeric(tbl[[x]][inds[[x]]]))
  tbl <- data.frame(t(do.call("rbind", tbl)))
  tbl <- tbl[2:nrow(tbl),]
  names(tbl) <- tbl_names
  
  tbl
}

tbl1 <- rbind(parse_table(txt[6:7], tbl_names = c("time", "inflow")),
              data.frame(time = seq(20, 36, by = 2), inflow = 0))
tbl2 <- parse_table(txt[9:10], tbl_names = c("storage", "discharge"))
tbl2$storage <- tbl2$storage * 1000000

```

data frome [here](https://www.caee.utexas.edu/prof/maidment/CE374KSpr13/Visual/HydrologicRouting.pptx)

```{r }
lt <- data.frame(time = seq(0, 210, by = 10),
                 inflow = c(seq(0, 360, by = 60), seq(320, 0, by = -40), rep(0, 6)))

qh <- data.frame(elevation = seq(0, 10, by = 0.5),
                 discharge = c(0, 3, 8, 17, 30, 43, 60, 78, 97, 117, 137, 156, 173, 190,
                               205, 218, 231, 242, 253, 264, 275))

res_ppt <- level_pool_routing(lt, qh, area = 43560, delta_t = 600, initial_outflow = 0, initial_storage = 0, linear.fit = FALSE)

plot(res_ppt$time, res_ppt$inflow)
lines(res_ppt$time, res_ppt$outflow)

res_pdf <- level_pool_routing(tbl1, tbl2, NA, delta_t = 7200, initial_outflow = 57, initial_storage = 75000000, linear.fit = FALSE)

plot(res_pdf$time, res_pdf$inflow)
lines(res_pdf$time, res_pdf$outflow)

```
